---
title: "Predictive models"
author: "Jim Costello"
date: "7/15/2026"
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
---


Note that you will need the following packages
* caret
* glmnet
* reshape2
* gplots


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("cluster")
library("factoextra") # plots and visualization
library("ggfortify") # plots and visualization
library("ggplot2") # plots and visualization
library("glmnet") # elastic net, ridge, lasso regression
library("caret") # package to aide with ML tasks
library("limma") # differential epxression
library("caTools")
library("gplots")  
library("reshape2")
library("randomForest")
library("readxl") # used to read .xlsx files
library("openxlsx") # used for data export as Excel workbooks
library("tidyverse") # data wrangling and ggplot2
library("rstatix") # pipe- and tidy-friendly statistical tests
library("ggrepel") # for labelling genes
library("ggforce") # for sina plots
library("tictoc") # timer
library("skimr") # data summary
library("broom") # tidying model objects
library("janitor") # data cleaning
library("patchwork") # assembling multiple plots
library("conflicted")
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("count", "dplyr")

setwd("/Users/costellj/Dropbox/Colorado/DS3/")
#detach("package:here", unload=TRUE) # run this to reset here()
library("here") # generates path to current project directory
here()
source(here("2026", "helper_functions.R")) # load helper functions

```

### Load Cytokine data and clinial information {#data}

The data we will use were collected from individuals with and without Down sydrome. The samples are from blood plasma and cytokine data were generated for each individual.

```{r, message=F}

# define the file paths to the input data
htp_meta_data_file <- here("2026/data", "HTP_Metadata_v0.5_Synapse.txt") 
htp_cytokines_data_file <- here("2026/data", "HTP_MSD_Cytokines_Synapse.txt") 

# Other parameters
# standard_colors <- c("Group1" = "#F8766D", "Group2" = "#00BFC4")
standard_colors <- c("Control" = "gray60", "T21" = "#009b4e")
out_file_prefix <- "linear_regression_htp_cytokines_v0.1_"
# End required parameters ###


# 1 Read in and inspect data ----
## 1.1 Read in meta data ----
htp_meta_data <- htp_meta_data_file |> 
  read_tsv() |> 
  mutate(
    Karyotype = fct_relevel(Karyotype, c("Control", "T21")), # convert to factor and set order
    Sex = fct_relevel(Sex, "Female"), # convert to factor and set order
    Sample_source_code = as_factor(Sample_source_code) # convert to factor - default is numerical order
  )
# inspect
htp_meta_data
htp_meta_data |> skimr::skim()
#
here("2026/data", "HTP_Metadata_v0.5_dictionary.txt") |> read_tsv()
#

## 1.2 Read in abundance data ----
htp_cytokines_data <- htp_cytokines_data_file |> 
  read_tsv()
  # janitor::clean_names(case = "none")
# inspect
htp_cytokines_data # 25,758 rows
htp_cytokines_data |> skimr::skim()
htp_cytokines_data |> distinct(Analyte) # 54 Analytes
htp_cytokines_data |> distinct(LabID) # 477 LabIDs
#
here("2026/data", "HTP_MSD_Cytokines_dictionary.txt") |> read_tsv()
#

## 1.3 Join meta data with data type 1 and data type 2 ----
htp_meta_cytokines_data <- htp_cytokines_data |> 
  inner_join(htp_meta_data, by="LabID")
# check number of rows returned !!!


# 2 Data exploration  ----
## 2.1 basic check of data distribution(s) ----
htp_meta_cytokines_data |> 
  filter(Analyte == "CRP") |> 
  ggplot(aes(Karyotype, log2(Value), color = Karyotype)) +
  geom_boxplot()
#

#create a data.frame of individuals by cytokines
cytokines_df <- as.data.frame(pivot_wider(htp_meta_cytokines_data, names_from = "Analyte", values_from = "Value", id_cols = "LabID"))
row.names(cytokines_df) <- cytokines_df$LabID
cytokines_df <- log2(cytokines_df[,-1])

# extract the annotations for each of the samples
pca_annos <- as.data.frame(htp_meta_data %>% filter(LabID %in% row.names(cytokines_df)))
row.names(pca_annos) <- pca_annos$LabID
pca_annos <- pca_annos[row.names(cytokines_df),]

# create a dataframe that will be used for plots that contains the cytokine and patient information.
cytokines_df_annos <- cbind(cytokines_df, pca_annos)

# PCA with HTP Cytokine data
pca <- prcomp(na.omit(cytokines_df), scale=T)

# remove outlier samples
hist(pca$x[,1], main ="PC1")
sort(pca$x[,1])
row.names(cytokines_df)[pca$x[,1] < -10]
cytokines_df <- cytokines_df[pca$x[,1] > -10,]
cytokines_df_annos <- cytokines_df_annos[pca$x[,1] > -10,]
dim(cytokines_df)
dim(cytokines_df_annos)
pca <- prcomp(na.omit(cytokines_df), scale=T)
autoplot(pca, data=cytokines_df_annos, col='Karyotype')



```

### Hierarchical Clustering {#hclust}

Note: Make sure you install the `hclust` package in R. Some of the code is taken from this nice [tutorial post](https://bradleyboehmke.github.io/HOML/hierarchical.html) on hierarchical clustering. 

```{r, message = F, warning = F}

# we will first perform agglomerative hierachical clustering (bottom up) using the agnes function in the cluster package

# test the various linkage methods to see which provides the strongest clusters
#define linkage methods
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

#function to compute agglomerative coefficient
ac <- function(x) {
  agnes(scale(cytokines_df), method = x)$ac
}

#calculate agglomerative coefficient for each clustering linkage method
sapply(m, ac)

#generate the agglomerative clustering and perform some visualizations
ac <- agnes(scale(cytokines_df), method="ward")
plot(as.hclust(ac), cex=0.3, main="Dendrogram for AGNES")
rect.hclust(ac, k = 4, border = 2:5)

# lets see how we can determine the appropriate cluster size

# The total within-cluster sum of square (wss) measures the compactness of the clustering and we want it to be as small as possible. 
p1 <- fviz_nbclust(scale(cytokines_df), FUN = hcut, method = "wss", 
                   k.max = 10) +
  ggtitle("(A) Elbow method")
# silhouette method determines how well each object lies within its cluster. A high average silhouette width indicates a good clustering.
p2 <- fviz_nbclust(scale(cytokines_df), FUN = hcut, method = "silhouette", 
                   k.max = 10) +
  ggtitle("(B) Silhouette method")
# The gap statistic compares the total intracluster variation for different values of k with their expected values under null reference distribution of the data (i.e. a distribution with no obvious clustering). The reference dataset is generated using Monte Carlo simulations of the sampling process.
p3 <- fviz_nbclust(scale(cytokines_df), FUN = hcut, method = "gap_stat", 
                   k.max = 10) +
  ggtitle("(C) Gap statistic")
gridExtra::grid.arrange(p1, p2, p3, nrow = 1)


# lets see where the clusters fall in PCA space
cytokines_df_annos$agnesclusters <- as.factor(cutree(ac, k=5))
autoplot(pca, data=cytokines_df_annos, col='agnesclusters')


# we will next perform divisive hierarchical clustering (top down) using the diana function in the cluster package
di <- diana(scale(cytokines_df))
di$dc
plot(as.hclust(di), cex=0.3, main="Dendrogram for DIANA")

cytokines_df_annos$dianaclusters <- as.factor(cutree(di, k=5))
autoplot(pca, data=cytokines_df_annos, col='dianaclusters')


```

### Linear Models {#glmnet}

Note: Make sure you install the `glmnet` package in R. We will use this for all examples of Ridge, Lasso and elastic net regression.

```{r, message = F, warning = F}

# we will predict on BMI, so remove individuals without any BMI information.
cytokines_df_annos$BMI <- cytokines_df_annos$Weight_kg/(cytokines_df_annos$Height_cm/100)^2
cytokines_df <- cytokines_df[!is.na(cytokines_df_annos$BMI),]
cytokines_df_annos <- cytokines_df_annos[!is.na(cytokines_df_annos$BMI),]

######## Organize the data for fitting linear models:
response = cytokines_df_annos$BMI

# Define cutoffs. We will use standard BMI scales to define normal, overweight, obese
normal.cutoff =  25  
obese.cutoff = 30

# Append response variable to matrix 
cytokines_df_annos$Response = ifelse(response >= obese.cutoff, 'obese',
                          ifelse(response <= normal.cutoff, 'normal', 'overweight'))

# Save group sizes for comparison
tab = table(cytokines_df_annos$Response)

# Check plot to make sure our cutoffs look reasonable
group.colors = ifelse(cytokines_df_annos$Response == 'obese', 'blue',
                      ifelse(cytokines_df_annos$Response == 'normal', 'red', 'black'))

plot(response, col = group.colors, pch = 16,
     main = 'BMI across the HTP cohort', 
     xlab = 'Individuals',
     ylab = 'BMI')
legend('topright', pch = 16, col = c('red', 'blue', 'black'),
       legend = c(paste0(sprintf('Normal (n = %d)', tab['normal'])), 
                  paste0(sprintf('Obese (n = %d)', tab['obese'])), 
                  paste0(sprintf('Overweight (n = %d)', tab['overweight']))))

# Filter out the overweight and make the two responses a factor
cytokines_df = cytokines_df[cytokines_df_annos$Response != 'overweight',]
cytokines_df_annos = cytokines_df_annos[cytokines_df_annos$Response != 'overweight',]
cytokines_df$Response <- factor(cytokines_df_annos$Response)
table(cytokines_df$Response)


# Define train and test sets
individuals = row.names(cytokines_df)
train.percent = .75
inTrain = individuals %in% sample(individuals, floor(train.percent*length(individuals)))
train.data = cytokines_df[inTrain,]
test.data = cytokines_df[!inTrain,]

# Make sure we have enough of each group in test set
table(test.data$Response)
table(train.data$Response)

```

#### Example 1 - Ridge regression

```{r message = F, warning = F}
# Make feature matrix and response vector
feature.matrix = as.matrix(train.data[,-ncol(train.data)])  # Exclude the last column (Response)
response.vector = train.data$Response

# Fit linear Ridge regression model
ridge.fit = glmnet(x = feature.matrix,      # features = all cytokines
                   y = response.vector,     # binary  response
                   family = 'binomial',     # we are doing binary classification
                   alpha = 0)               # alpha = 0 is the Ridge penalty

# Use the fit model to predict on the testing data
testing.matrix= as.matrix(test.data[,-ncol(test.data)]) # Exclude the last column (Response)
testing.response = test.data$Response
ridge.preds = predict(ridge.fit, newx = testing.matrix, type = 'class')

# Take a look at how our model did
table(Predicted_Group = ridge.preds[,ncol(ridge.preds)], 
      Actual_Group = test.data$Response)

ReportPerfMetrics = function(predicted.labels, true.labels, pos.class){
  # Calculate the accuracy, precision and recall for two-class prediction
  tp = sum(true.labels == pos.class & predicted.labels == pos.class)
  fp = sum(true.labels != pos.class & predicted.labels == pos.class)
  tn = sum(true.labels != pos.class & predicted.labels != pos.class)
  fn = sum(true.labels == pos.class & predicted.labels != pos.class)
  n = tp + fp + tn + fn
  
  accuracy = (tp + tn)/n
  precision = tp/(tp + fp)
  recall = tp/(tp + fn)
  
  return(list(Accuracy = accuracy, Precision = precision, Recall = recall))
}

ridge.metrics = ReportPerfMetrics(ridge.preds[,ncol(ridge.preds)], test.data$Response, 'obese')
ridge.metrics

```

#### Example 2 - Lasso regression

```{r, message = F, warning = F}

# Fit linear Lasso regression model
lasso.fit = glmnet(x = feature.matrix,      # features = all cytokines
                   y = response.vector,     # binary response
                   family = 'binomial',     # we are doing binary classification
                   alpha = 1)               # alpha = 1 is the Lasso penalty

# Use the fit model to predict on the testing data
lasso.preds = predict(lasso.fit, newx = testing.matrix, type = 'class')

# Take a look at how our model did
table(Predicted_Group = lasso.preds[,ncol(lasso.preds)], 
      Actual_Group = test.data$Response)
lasso.metrics = ReportPerfMetrics(lasso.preds[,ncol(lasso.preds)], test.data$Response, 'obese')
lasso.metrics

#lets compare lasso and ridge coefficients
plot(lasso.fit, xvar="lambda", label=T)
plot(ridge.fit, xvar = "lambda", label = TRUE)

```

#### Example 3 - Elastic net regression with cross-validation

```{r, message = F, warning = F}

# Fit elastic net model
elastic.fit = cv.glmnet(x = feature.matrix,      # features = all cytokines
                   y = response.vector,          # binary response
                   family = 'binomial',          # we are doing binary classification
                   nfolds = 5,
                   type.measure = 'auc',
                   alpha = 0.5)

# Use the fit model to predict on the testing data
elastic.preds = predict(elastic.fit, newx = testing.matrix, type = 'class')

# Take a look at how our model did
elastic.metrics = ReportPerfMetrics(elastic.preds[,ncol(elastic.preds)], test.data$Response, 'obese')
table(Predicted_Group = elastic.preds[,ncol(elastic.preds)], 
      Actual_Group = test.data$Response)

# Plot the cross-validation curve, and upper and lower standard deviation curves, as a function of the lambda values used
plot(elastic.fit)

```
  
### Sparse Linear Regression Evaluation {#evaluation}

For discussion on the alpha and lambda parameters, check out this [review of `glmnet` and elastic net regression](http://www.moseslab.csb.utoronto.ca/alan/glmnet_presentation.pdf).

```{r, message = F, warning = F}
# Compare the 3 examples, alpha = 0, alpha = 1, and alpha = 0.5
results = cbind(Ridge = ridge.metrics, Lasso = lasso.metrics, ElasticNet = elastic.metrics)
results

######### Find optimal alpha and lambda with caret cross-validation
# Construct object for holding training parameters
my.train.control = trainControl(method = "repeatedcv", number = 5, returnResamp = "all")

# Train an elastic net model with varying alpha and lambda
model = train(Response ~ ., data = train.data, 
               method = "glmnet",                                        # Fit an elastic net model
               metric = "Accuracy",                                      # Use accuracy as the loss for cross validation
               tuneGrid = expand.grid(.alpha = seq(0, .5, by = .05),     # Try these alpha values
                                      .lambda = seq(0, 1, by = .05)),    # And these lambda values
               trControl = my.train.control)
model

# Reshape the data into a matrix for making a heatmap
model.results = model$results
model.cast = dcast(model.results, alpha ~ lambda, value.var = 'Accuracy')
row.names(model.cast) = model.cast$alpha
model.cast$alpha = NULL

# Make a heatmap of the alphas and lambdas
my.palette <- colorRampPalette(c("blue", "red"))(n = 100)
heatmap.2(as.matrix(model.cast), 
          col = my.palette,           # Define heatmap colors
          Rowv = F, Colv = F,         # Don't cluster/reorder rows or columns
          dendrogram = 'none',        # Don't plot dendrogram
          trace = 'none',             # Don't draw trace lines
          density.info = 'none',      # Don't draw histogram on color key
          key.xlab = 'ROC',           # Label the color key
          main = 'Optimizing alpha and lambda values',
          xlab = 'Lambda values',
          ylab = 'Alpha values')             

```

### Random Forest {#randomforest}

This is a nice reference for the basics on the [Random Forest](https://www.simplilearn.com/tutorials/data-science-tutorial/random-forest-in-r). You can also review the [Statquest](https://www.youtube.com/watch?v=J4Wdy0Wc_xQ&t=1s) on Random Forests.

```{r, message = F, warning = F}

# Fitting Random Forest to the train dataset 
set.seed(120)  # Setting seed 
classifier_RF = randomForest(x = feature.matrix, 
                             y = response.vector, 
                             ntree = 500) 
classifier_RF 
  
# Predicting the Test set results 
y_pred = predict(classifier_RF, newdata = testing.matrix) 
  
# Confusion Matrix 
confusion_mtx = table(testing.response, y_pred) 
confusion_mtx 
  
# Plotting model 
plot(classifier_RF) 
legend("topright", legend=colnames(classifier_RF$err.rate), fill=c("black", "red", "green"))
  
# Importance plot 
importance(classifier_RF) 
  
# Variable importance plot 
varImpPlot(classifier_RF) 
```

### Session Information {#session}

```{r, message = F}
sessionInfo()
```


